The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf

### 2015-03-19 oil_prices_rr.tex
work on mc

%\doublespacing % from package setspacs

@@

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/top10_production_print.png}
	\caption{The production profile of the 10 largest oil fields on the Norwegian Continental Shelf.  Production tends to be correlated across fields}
	\label{top10_production}	
	\end{figure}

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/oil_decline_print.png}
	\caption{The production profile for the entire Norwegian Continental Shelf is bell-shaped, reflecting the correlated production profiles of the fields.  With oil prices that are autocorrelated, }
	\label{oil_decline}
\end{figure}

As a solution I use a semi-parametric model within the Generalized Additive Model frameworks of \cite{hastie_generalized_1990}.  Here I use a two-dimensional smoothed spline function to account for the general non-parametric shape of the production profile while allowing price to enter the equation linearly.  The coefficient of price can then be interpreted as the average effect of price on production over the entire production profile.

@@

\begin{figure}
\includegraphics[width=1.2\textwidth]{figures/north_sea_reserves_print.png}
\caption{Oil fields in the Norwegian territorial North Sea.  The largest oil fields tended to be discovered earliest, while newer finds tend to be smaller.  An exception is the large Johan Sverdrup field, which is expected to begin producing in 2019.}
\label{north_sea_reserves}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1.2\textwidth]{figures/norwegian_sea_reserves_print.png}
\caption{Oil fields in the Norwegian Sea.  Production from the Norwegian Sea has generally been dissapointing compared with expectations when the area was opened to exploration in the early 1980's.}
\label{norwegian_sea_reserves}
\end{figure}

@@

A cleaned data set and the full code for the analysis are available upon request. I use the R statistical programming package for all the analysis in this article \citep{r_core_team_r:_2013}.  I use the R packages ggplot2 and ggmap for plotting \citep{wickham_ggplot2:_2009, kahle_ggmap:_2013}, plyr for data manipulation and cleaning \citep{wickham_split-apply-combine_2011}, texreg for table formatting \citep{leifeld_texreg:_2013} and mgcv for implementation of the Generalized Additive Models \citep{wood_fast_2011}.

@@

@@

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/gam_postpeak_print.png}
	\caption{The estimated coefficients on the price terms in a model of oil field production in depleting fields.  Significant coefficients are only estimated on the concurrent price and eight lag, however the significance of the estimated coefficient on the eighth lag is not robust to specification}
	\label{gam_postpeak_print}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/gam_prepeak_print.png}
	\caption{The estimated coefficients on the price terms in a model of fields in the build-out phase.  For large fields, positive coefficients are estimated at the 6th, 7th and 8th lag, but the estimates are imprecise and not statistically significant.  Significant positive coefficients are estimated for small fields at the fifth and eighth lag.  In a pooled model, significant positive coefficients are found at the fourth through eighth lags.}
	\label{gam_prepeak_print}
\end{figure}

@@

@@

% \end{spacing}
% \end{document} for shorter version with no appendix
\FloatBarrier

\appendix

\section{Appendix: Tables of full results and robustness checks}

\begin{table}
\begin{center}
\begin{tabular}{l c c c c c c }
\hline
	&\multicolumn{2}{|c|}{6 lags} & \multicolumn{2}{c}{8 lags} &\multicolumn{2}{c}{4-8 lags} \\
                                     & Small & Large & Small  & Large & Small & Large\\
\hline
(Intercept)                          & $-1.46^{***}$ & $-0.15$       & $-1.58^{***}$ & $-0.25$       & $-1.54^{***}$ & $-1.54^{***}$ \\
                                     & $(0.37)$      & $(0.50)$      & $(0.38)$      & $(0.51)$      & $(0.37)$      & $(0.37)$      \\
oilpricereal                       & $0.00$        & $-0.02$       & $0.00$        & $-0.01$       &               &               \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
oilpricereall1                    & $0.01$        & $-0.01$       & $0.01$        & $0.01$        &               &               \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
oilpricereall2                    & $-0.02$       & $0.01$        & $-0.01$       & $0.00$        &               &               \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
oilpricereall3                    & $-0.02$       & $0.01$        & $-0.03^{*}$   & $0.00$        &               &               \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      &               &               \\
oilpricereall4                    & $0.00$        & $0.03^{*}$    & $0.00$        & $0.03^{*}$    & $-0.01$       & $-0.01$       \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall5                    & $-0.01$       & $0.01$        & $-0.01$       & $0.01$        & $-0.02$       & $-0.02$       \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall6                    & $0.02^{*}$    & $0.03^{*}$    & $0.02$        & $0.01$        & $0.02^{*}$    & $0.02^{*}$    \\
                                     & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
EDF: s(timetopeak,recoverableoil) & $28.92^{***}$ & $26.89^{***}$ & $28.91^{***}$ & $26.89^{***}$ & $28.91^{***}$ & $28.91^{***}$ \\
                                     & $(29.00)$     & $(27.00)$     & $(29.00)$     & $(27.00)$     & $(29.00)$     & $(29.00)$     \\
EDF: s(peaktoend,recoverableoil)  & $17.75^{***}$ & $10.71^{***}$ & $17.58^{***}$ & $11.41^{***}$ & $17.47^{***}$ & $17.47^{***}$ \\
                                     & $(28.00)$     & $(28.00)$     & $(28.00)$     & $(28.00)$     & $(28.00)$     & $(28.00)$     \\
oilpricereall7                    &               &               & $-0.01$       & $0.02^{*}$    & $-0.01$       & $-0.01$       \\
                                     &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall8                    &               &               & $0.04^{***}$  & $0.03^{**}$   & $0.03^{***}$  & $0.03^{***}$  \\
                                     &               &               & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
\hline
AIC                                  &               & 1211.33       &               & 1186.97       &               &               \\
BIC                                  &               & 1377.97       &               & 1363.28       &               &               \\
Log Likelihood                       &               & -559.07       &               & -544.18       &               &               \\
Deviance                             & 12319.62      & 301090.57     & 12104.52      & 268975.69     & 12347.62      & 12347.62      \\
Deviance explained                   & 0.84          & 0.96          & 0.85          & 0.96          & 0.84          & 0.84          \\
Dispersion                           & 15.17         & 1378.61       & 14.93         & 1247.02       & 15.16         & 15.16         \\
R$^2$                                & 0.82          & 0.95          & 0.83          & 0.95          & 0.82          & 0.82          \\
GCV score                            & 16.19         & 1666.44       & 15.98         & 1526.28       & 16.13         & 16.13         \\
Num. obs.                            & 865           & 264           & 865           & 264           & 865           & 865           \\
Num. smooth terms                    & 2             & 2             & 2             & 2             & 2             & 2             \\
\hline
\multicolumn{7}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Full results for model of field production with price, seperate estimation of small and large fields.}
\label{table:unpooled}
\end{center}
\end{table}

%table two pooled
\begin{table}
\begin{center}
\begin{tabular}{l c c }
\hline
                                     & Pooled Model & Insignificant terms dropped \\
\hline
(Intercept)                          & $-1465.35^{***}$ & $-1500.51^{***}$ \\
                                     & $(65.21)$        & $(66.28)$        \\
oilpricereal                       & $-0.01$          &                  \\
                                     & $(0.01)$         &                  \\
largefieldsmall                     & $-2.32^{***}$    & $-1.19^{***}$    \\
                                     & $(0.27)$         & $(0.10)$         \\
oilpricereall1                    & $0.02^{*}$       &                  \\
                                     & $(0.01)$         &                  \\
oilpricereall2                    & $0.01$           &                  \\
                                     & $(0.01)$         &                  \\
oilpricereall3                    & $0.00$           &                  \\
                                     & $(0.01)$         &                  \\
oilpricereall4                    & $0.03^{***}$     & $0.03^{***}$     \\
                                     & $(0.01)$         & $(0.01)$         \\
oilpricereall5                    & $0.01$           & $0.01$           \\
                                     & $(0.01)$         & $(0.01)$         \\
oilpricereall6                    & $0.02^{***}$     & $0.03^{***}$     \\
                                     & $(0.01)$         & $(0.01)$         \\
oilpricereall7                    & $0.02^{**}$      & $0.02^{***}$     \\
                                     & $(0.01)$         & $(0.01)$         \\
oilpricereall8                    & $0.03^{***}$     & $0.03^{***}$     \\
                                     & $(0.00)$         & $(0.00)$         \\
oilpricereal:largefieldsmall      & $-0.03$          &                  \\
                                     & $(0.06)$         &                  \\
largefieldsmall:oilpricereall1   & $0.03$           &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall2   & $-0.05$          &                  \\
                                     & $(0.08)$         &                  \\
largefieldsmall:oilpricereall3   & $0.03$           &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall4   & $0.03$           &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall5   & $0.04$           &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall6   & $0.18^{*}$       &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall7   & $0.06$           &                  \\
                                     & $(0.07)$         &                  \\
largefieldsmall:oilpricereall8   & $0.07$           &                  \\
                                     & $(0.07)$         &                  \\
EDF: s(timetopeak,recoverableoil) & $28.99^{***}$    & $28.99^{***}$    \\
                                     & $(29.00)$        & $(29.00)$        \\
EDF: s(peaktoend,recoverableoil)  & $27.98^{***}$    & $26.99^{***}$    \\
                                     & $(28.00)$        & $(27.00)$        \\
\hline
AIC                                  &                  &                  \\
BIC                                  &                  &                  \\
Log Likelihood                       &                  &                  \\
Deviance                             & 371079.07        & 396494.49        \\
Deviance explained                   & 0.96             & 0.96             \\
Dispersion                           & 352.06           & 371.24           \\
R$^2$                                & 0.96             & 0.95             \\
GCV score                            & 377.77           & 393.13           \\
Num. obs.                            & 1129             & 1129             \\
Num. smooth terms                    & 2                & 2                \\
\hline
\multicolumn{3}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Full results for pooled model of field production with price terms.}
\label{table:pooled}
\end{center}
\end{table}

%investment tables

\begin{table}
\begin{center}
\begin{tabular}{l c c c }
\hline
                                     & Pooled & Small Fields & Large Fields \\
\hline
(Intercept)                          & $-199.84^{***}$ & $7.03^{***}$   & $7.64^{***}$    \\
                                     & $(44.00)$       & $(0.36)$       & $(0.24)$        \\
yearprod                            & $-0.06^{***}$   & $-0.26^{***}$  & $-0.06^{***}$   \\
                                     & $(0.01)$        & $(0.02)$       & $(0.01)$        \\
oilpricereal                       & $0.08^{***}$    & $-0.03^{**}$   & $0.04$          \\
                                     & $(0.01)$        & $(0.01)$       & $(0.03)$        \\
largefieldsmall                     & $1.80^{***}$    &                &                 \\
                                     & $(0.52)$        &                &                 \\
oilpricereall1                    & $-0.02$         & $0.00$         & $0.00$          \\
                                     & $(0.01)$        & $(0.01)$       & $(0.03)$        \\
oilpricereall2                    & $0.06^{***}$    & $0.05^{***}$   & $0.09^{***}$    \\
                                     & $(0.01)$        & $(0.01)$       & $(0.03)$        \\
oilpricereall3                    & $0.16^{***}$    & $0.25^{***}$   & $0.19^{***}$    \\
                                     & $(0.01)$        & $(0.03)$       & $(0.03)$        \\
oilpricereall4                    & $0.00$          & $-0.08^{**}$   & $-0.03$         \\
                                     & $(0.01)$        & $(0.03)$       & $(0.03)$        \\
oilpricereall5                    & $-0.02$         & $-0.09^{**}$   & $-0.07^{*}$     \\
                                     & $(0.02)$        & $(0.03)$       & $(0.03)$        \\
oilpricereall6                    & $-0.09^{***}$   & $-0.10^{**}$   & $-0.04$         \\
                                     & $(0.02)$        & $(0.04)$       & $(0.04)$        \\
oilpricereall7                    & $-0.11^{***}$   & $-0.26^{***}$  & $-0.09^{*}$     \\
                                     & $(0.02)$        & $(0.04)$       & $(0.04)$        \\
oilpricereall8                    & $-0.10^{***}$   & $-0.19^{***}$  & $-0.12^{***}$   \\
                                     & $(0.01)$        & $(0.03)$       & $(0.03)$        \\
oilpricereal:smallfield      & $-0.11^{*}$     &                &                 \\
                                     & $(0.05)$        &                &                 \\
largefieldsmall:oilpricereall1   & $0.02$          &                &                 \\
                                     & $(0.05)$        &                &                 \\
largefieldsmall:oilpricereall2   & $0.01$          &                &                 \\
                                     & $(0.05)$        &                &                 \\
largefieldsmall:oilpricereall3   & $0.03$          &                &                 \\
                                     & $(0.06)$        &                &                 \\
EDF: s(timetopeak,recoverableoil) & $28.99^{***}$   & $26.92^{***}$  & $28.79^{***}$   \\
                                     & $(29.00)$       & $(28.11)$      & $(28.97)$       \\
EDF: s(peaktoend,recoverableoil)  & $16.90^{***}$   & $4.83^{*}$     & $5.39^{**}$     \\
                                     & $(18.82)$       & $(6.40)$       & $(7.44)$        \\
\hline
AIC                                  &                 &                & 4747.87         \\
BIC                                  &                 &                & 4912.98         \\
Log Likelihood                       &                 &                & -2327.76        \\
Deviance                             & 257371493567.88 & 10630461533.61 & 198564735228.80 \\
Deviance explained                   & 0.82            & 0.81           & 0.84            \\
Dispersion                           & 243467395.17    & 13087671.29    & 907398412.98    \\
R$^2$                                & 0.80            & 0.79           & 0.81            \\
GCV score                            & 257721863.13    & 13776495.42    & 1094706852.34   \\
Num. obs.                            & 1117            & 853            & 264             \\
Num. smooth terms                    & 2               & 2              & 2               \\
\hline
\multicolumn{4}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Full results for model of field investment with price terms.}
\label{table:investment}
\end{center}
\end{table}

% Table post-peak

\begin{table}
\begin{center}
\begin{tabular}{l c c c c }
\hline
&\multicolumn{2}{c}{Recoverable Oil} & \multicolumn{2}{c}{Max Production} \\
                                    & Large & Small & Large & Small \\
\hline
(Intercept)                         & $1.54^{***}$  & $-0.31^{*}$   & $2.25^{***}$  & $-1.61^{***}$ \\
                                    & $(0.06)$      & $(0.14)$      & $(0.08)$      & $(0.25)$      \\
oilpricereal                      & $0.01$        & $0.03^{***}$  & $-0.02^{*}$   & $0.02^{*}$    \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall1                   & $0.02$        & $-0.01$       & $-0.01$       & $0.00$        \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall2                   & $0.00$        & $-0.01$       & $-0.02$       & $-0.01$       \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall3                   & $0.02$        & $-0.03^{**}$  & $0.00$        & $-0.02^{*}$   \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall4                   & $0.03$        & $0.00$        & $0.01$        & $0.00$        \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall5                   & $-0.01$       & $-0.03^{*}$   & $-0.01$       & $-0.02$       \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall6                   & $0.01$        & $0.01$        & $0.00$        & $0.00$        \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall7                   & $0.01$        & $-0.03$       & $0.02$        & $-0.01$       \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
oilpricereall8                   & $0.01$        & $0.06^{***}$  & $-0.01$       & $0.04^{***}$  \\
                                    & $(0.01)$      & $(0.01)$      & $(0.01)$      & $(0.01)$      \\
EDF: s(peaktoend,recoverableoil) & $27.63^{***}$ & $23.05^{***}$ &               &               \\
                                    & $(29.00)$     & $(29.00)$     &               &               \\
EDF: s(peaktoend,maxprod)        &               &               & $23.82^{***}$ & $28.82^{***}$ \\
                                    &               &               & $(29.00)$     & $(29.00)$     \\
\hline
AIC                                 & 571.09        & 1543.57       & 490.28        & 1421.72       \\
BIC                                 & 690.61        & 1696.16       & 598.01        & 1600.20       \\
Log Likelihood                      & -246.92       & -737.74       & -210.31       & -671.04       \\
Deviance                            & 49634.76      & 6868.08       & 31676.52      & 5599.00       \\
Deviance explained                  & 0.99          & 0.87          & 0.99          & 0.90          \\
Dispersion                          & 395.91        & 11.08         & 245.22        & 9.12          \\
R$^2$                               & 0.98          & 0.86          & 0.99          & 0.88          \\
GCV score                           & 514.74        & 11.67         & 309.43        & 9.69          \\
Num. obs.                           & 163           & 653           & 163           & 653           \\
Num. smooth terms                   & 1             & 1             & 1             & 1             \\
\hline
\multicolumn{5}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Full results for model of field production with price terms in depletion phase.  Using both estimate of total recoverable oil and max production to control for field size}
\label{table:postpeak}
\end{center}
\end{table}

% prepeak

\begin{table}
\begin{center}
\begin{tabular}{l c c c }
\hline
                                     & Small Fields & Large Fields & Pooled \\
\hline
(Intercept)                          & $-0.43$       & $1.26$        & $-0.91$       \\
                                     & $(0.29)$      & $(2.28)$      & $(0.68)$      \\
oilpricereal                       & $0.01$        & $-0.04$       & $0.02$        \\
                                     & $(0.03)$      & $(0.07)$      & $(0.04)$      \\
oilpricereall1                    & $0.03$        & $-0.06$       & $0.10^{**}$   \\
                                     & $(0.04)$      & $(0.09)$      & $(0.03)$      \\
oilpricereall2                    & $-0.01$       & $-0.11$       & $0.05^{*}$    \\
                                     & $(0.03)$      & $(0.11)$      & $(0.02)$      \\
oilpricereall3                    & $-0.07^{*}$   & $-0.11$       & $-0.02$       \\
                                     & $(0.03)$      & $(0.13)$      & $(0.02)$      \\
oilpricereall4                    & $0.06$        & $-0.02$       & $0.07^{***}$  \\
                                     & $(0.03)$      & $(0.12)$      & $(0.02)$      \\
oilpricereall5                    & $0.09^{**}$   & $0.01$        & $0.06^{***}$  \\
                                     & $(0.03)$      & $(0.10)$      & $(0.02)$      \\
oilpricereall6                    & $-0.01$       & $0.07$        & $0.04^{*}$    \\
                                     & $(0.03)$      & $(0.07)$      & $(0.02)$      \\
oilpricereall7                    & $0.01$        & $0.10$        & $0.05^{**}$   \\
                                     & $(0.03)$      & $(0.06)$      & $(0.02)$      \\
oilpricereall8                    & $0.07^{**}$   & $0.06$        & $0.05^{**}$   \\
                                     & $(0.02)$      & $(0.04)$      & $(0.02)$      \\
EDF: s(time_to_peak,recoverable_oil) & $16.91^{***}$ & $28.37^{***}$ & $20.51^{***}$ \\
                                     & $(29.00)$     & $(29.00)$     & $(29.00)$     \\
\hline
AIC                                  & 551.59        & 330.20        & 1534.58       \\
BIC                                  & 638.09        & 421.97        & 1644.24       \\
Log Likelihood                       & -247.89       & -125.73       & -735.78       \\
Deviance                             & 3949.57       & 34833.67      & 300487.73     \\
Deviance explained                   & 0.79          & 0.99          & 0.92          \\
Dispersion                           & 28.81         & 925.75        & 1434.35       \\
R$^2$                                & 0.74          & 0.98          & 0.91          \\
GCV score                            & 34.46         & 1869.82       & 1643.21       \\
Num. obs.                            & 164           & 76            & 240           \\
Num. smooth terms                    & 1             & 1             & 1             \\
\hline
\multicolumn{4}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Full results for model of field production with price terms in depletion phase.}
\label{table:prepeak}
\end{center}
\end{table}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/gam_under_2d_plot_1.png}
	\caption{Plot of estimated function of the production profile of small fields, time-to-peak.}
	\label{chart:gam_under_2d_plot_1}
\end{figure}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/gam_under_2d_plot_2.png}
	\caption{Plot of estimated function of the production profile of small fields, peak-to-end.}
	\label{chart:gam_under_2d_plot_2}
\end{figure}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/gam_over_2d_plot_1.png}
	\caption{Plot of estimated function of the production profile of large fields, time-to-peak.}
	\label{chart:gam_over_2d_plot_1}
\end{figure}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/gam_over_2d_plot_2.png}
	\caption{Plot of estimated function of the production profile of large fields, peak-to-end.}
	\label{chart:gam_over_2d_plot_2}
\end{figure}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/qq_gam_price_over_2d.png}
	\caption{QQplot of model residuals against theoretical quantiles of normal distribution.  A Gaussian distribution appears to be adequate.}
	\label{chart:qqplot}
\end{figure}

%Fitted vs. residuals
\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/under_fitted.png}
	\caption{Fitted verses actual production for small fields.}
	\label{chart:under_fitted}
\end{figure}

\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/over_fitted.png}
	\caption{Fitted verses actual production for larger fields.}
	\label{chart:under_fitted}
\end{figure}

% Check to see if overfitting of model - change gamma parameter to 1.4
\begin{table}
\begin{center}
\begin{tabular}{l c c }
\hline
                                     & Small Fields & Large Fields \\
\hline
(Intercept)                          & $-0.24^{***}$ & $1.88^{***}$  \\
                                     & $(0.07)$      & $(0.11)$      \\
oilpricereal                       & $0.02$        & $-0.04$       \\
                                     & $(0.01)$      & $(0.02)$      \\
oilpricereall1                    & $-0.01$       & $-0.02$       \\
                                     & $(0.01)$      & $(0.03)$      \\
oilpricereall2                    & $-0.01$       & $0.00$        \\
                                     & $(0.01)$      & $(0.03)$      \\
oilpricereall3                    & $-0.02$       & $0.00$        \\
                                     & $(0.01)$      & $(0.03)$      \\
oilpricereall4                    & $0.02$        & $0.02$        \\
                                     & $(0.01)$      & $(0.03)$      \\
oilpricereall5                    & $0.01$        & $-0.01$       \\
                                     & $(0.02)$      & $(0.03)$      \\
oilpricereall6                    & $-0.04^{*}$   & $0.09^{**}$   \\
                                     & $(0.02)$      & $(0.03)$      \\
EDF: s(time_to_peak,recoverable_oil) & $23.93^{***}$ & $28.58^{***}$ \\
                                     & $(29.00)$     & $(29.00)$     \\
EDF: s(peak_to_end,recoverable_oil)  & $25.91^{***}$ & $6.10^{***}$  \\
                                     & $(28.00)$     & $(28.00)$     \\
\hline
AIC                                  & 59890.49      & 441209.87     \\
BIC                                  & 60170.73      & 441366.08     \\
Log Likelihood                       & -29886.40     & -220561.25    \\
Deviance                             & 4653.55       & 12307.60      \\
Deviance explained                   & 0.80          & 0.83          \\
Dispersion                           & 5.75          & 55.61         \\
R$^2$                                & 0.77          & 0.89          \\
GCV score                            & 6.16          & 66.34         \\
Num. obs.                            & 865           & 264           \\
Num. smooth terms                    & 2             & 2             \\
\hline
\multicolumn{3}{l}{\scriptsize{\textsuperscript{***}$p<0.001$, 
  \textsuperscript{**}$p<0.01$, 
  \textsuperscript{*}$p<0.05$}}
\end{tabular}
\caption{Results when changing the $\gamma$ parameter that controls smoothness to 1.4 in case of overfitting, though this does not signficantly change results}
\label{table:gamma_check}
\end{center}
\end{table}

% max production vs. total fields
\begin{figure}
	\includegraphics[width=.8\textwidth]{figures/Max_prod_vs_TRO_print.png}
	\caption{Comparison of fits between using max production and estimate of total reserves to control for field size.}
	\label{chart:Max_prod_vs_TRO_print}
\end{figure}

\FloatBarrier

\section{Appendix: Monte Carlo Experiment}

The data set used in this paper has a complex structure.  The dependent variable - field level oil production - has a non-linear autocorrelated structure.  The production profiles are also correlated with each other - leading to a bell-shaped total production curve.  In addition the exogenous variable of interest - oil prices - are autocorrelated and non-stationary.    In such a setting, standard rules for inference may not hold.

In this appendix I instead present a Monte Carlo simulation experiment.  Generating artificial data that is similar in character to the data used, I first show how failing to properly control for the production profile can lead to a biased estimation of the effect of oil prices.  I then show how using a generalized additive model can lead to better estimates.

I use the R statistical programming language \citet{r_core_team_r:_2013} for the analysis.  For illustration, I include code snippets in this document.  The full code for the analysis is available upon request. The figures shown are created using the r package ggplot2 \citet{wickham_ggplot2:_2009}.

\subsection{Generating Data}

The first step is generate a set of fields from a size distribution.  In the simulation, I generate 77 fields from a exponential normal distribution with mean 2.3 and standard deviation of 1.5.  This is similar to the distribution of the actual 77 productive Norwegian fields. 
\begin{verbatim}
field_size<-round(exp(rnorm(77, mean=2.3, sd=1.5)), digits=1)
\end{verbatim}

Using the generated field sizes as an input, I then create a function to generate starting times.  The goal here is to mimic what is called creaming in the exploration industry - the tendency of finding and starting production on larger fields first. For fields that are smaller than 10 million SM3, I generate a starting time from a uniform distribution with endpoints 1975 and 2008.  For larger fields, I generate a starting time from a normal distribution with standard deviation of 10 and a mean as in formula 1:

\begin{equation}
mean_start(size) = trunc(1973 + (max + 300)/size)
\end{equation}

\begin{verbatim}
gen_year<-function(size, maxsize){
	#let small fields be distributed uniformly from 1975 to 2013
	if(size<10){
		year<-trunc(runif(1, 1975, 2008))	
	} else{	#while large fields should be more common earlier on
		range<-FALSE
		while(range==FALSE){
			year<-trunc(rnorm(1,mean=(1973+(maxsize+300)/(size+300)), sd=10))
			ifelse(year>=1970 & year<=2013, range<-TRUE, range<-FALSE)
			}	
		}
return(year)	
}
\end{verbatim}

Generating the starting times in this way was essentially ad-hoc.  However the method is sufficient since the purpose of this exercise is not to model the underlying physical and economic processes, but rather to generate data that shares key similarities with the actual data.

In the simulation I use the actual Brent oil price series shown in figure \ref{oil_price_series}.  This series can be shown to be non-stationary.  In order to get reliable inference in a simpler time-series setting the solution might be to take the first difference of the series.  This is not possible in this setting since it is the actual levels of oil prices that are important.  Intuitively, identification comes from comparison of fields of similar size and at similar points in their production lifetime but at periods at differing oil prices.  This variation would be discarded by purely looking at first differences.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/oil_price_series.png}
	\caption{The Real Price of Brent Traded Oil from 1960 to 2013 in 2010 U.S. Dollars.}
	\label{oil_price_series}	
\end{figure}

So far, all the simulated data has been outside the Monte Carlo function.  The data is treated as being fixed.Within the Monte-Carlo loop I create a function that takes as input a fields size, start date as well as the price series and generates a production profile.  The code for this function is below.

\begin{verbatim}
gen_production<-function(field, prices){
	size<-field[1]
	start<-field[2]
	name<-field[3]

	names(prices)<-c("year", "price")
	t<-trunc(sqrt(size)) + 3  #generate production time
	prod_time<- -t:(t)
	year=start:(start+2*t-1)

	#use a cumulative logistic function to represent shape of function
	cum_production = 1/(1+exp(-(prod_time)/3))*size

	#take difference to create production per year
	prod_shape = diff(cum_production)
	production<-data.frame(year, prod_shape)

	#Formula: log(production) = f`(time) + beta*log(price) + epsilon

	production$prod<-production$prod_shape*
	exp(beta0*production$price*rlnorm(length(production$price), meanlog=0, sdlog=.05))

	return(production)
}
\end{verbatim}

First, the production time of the field is created as a deterministic function of field size.  I then model cumulative production from the field over time as a logistic function as in equation \ref{cumProd_equation}.  This series is then first-differenced to create the general production profile.

\begin{equation}
% 1/(1+exp(-(prod_time)/3))*size
cumProd=\frac{size}{1+exp(\frac{-prodTime_t}{3})}
\label{cumProd_equation}

\end{equation}

I multiply the production profile series with an exponential price term and a coefficient representing the effect of price as well as a random component with mean 0 and standard deviation .05 drawn from a log-normal distribution.  Taking the logarithm of both sides, the production series for a given field can be written as equation \ref{production_equation}.  Importantly, the term $\beta$ represents the true effect of price on production.

\begin{equation}
log(production_t)=log(prodShape_t) + \beta price_t + \epsilon_t
\label{production_equation}
\end{equation}

Within the Monte-Carlo loop, production data for each field is generated as above.  Thus the variation in each run of the Monte Carlo simulation is the random term $\epsilon_t$ from equation \ref{production_equation}.

An example of the data generated can be seen in figure \ref{simulated_production}.  By inspection, it appears reasonably similar to the data on actual oil production on the Norwegian Continental Shelf.

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/simulated_production.png}
	\caption{Simulated production of 77 oil fields}
	\label{simulated_production}	
\end{figure}

\subsection{Estimation of Coefficient}

Within the Monte Carlo function, a regression is run using the gam() function from the package mgcv \citet{wood_generalized_2006}.  This will handle both estimation of linear models or generalized additive models, depending on what formula is passed to the Monte Carlo function.

\begin{verbatim}
gam_sim<-gam(formula,
	family=gaussian(link=log), weights=size, data=sim_fields)
summary(gam_sim)
\end{verbatim}

Outside of the Monte Carlo function, First I specify a linear regression formula with a 3rd degree polynomial terms to control for the production profile as below.

\begin{verbatim}
formula_0=formula(prod~time_to_peak + time_to_peak_sq + time_to_peak_cu + 
	peak_to_end + peak_to_end_sq + peak_to_end_cu +size + price)
\end{verbatim}

I then replicate the data 1000 times, letting the $\beta$ term equal zero, that is having no effect of prices on oil production.  The formula used is shown below, where $gam_mc()$ is the Monte Carlo function.

\begin{verbatim}
gamm_mc_0<-replicate(1000, gam_mc(beta=0, formula=formula_0, use_true_prices=TRUE))
\end{verbatim}

The above line returns 1000 estimates of $\beta$ from the linear regression specification.  Figure \ref{lin_model_price_mc} shows the results in the form of a histogram.

>>>>

\doublespacing % from package setspacs

@@

As a solution I use a semi-parametric model within the Generalized Additive Model frameworks of \cite{hastie_generalized_1990}.  Here I use a cubic spline function to account for the general non-parametric shape of the production profile while allowing price to enter the equation linearly while also attempting to control for the effects of varying production costs and the effects of technological change and other time-varying effects.

The results suggest that little to no concurrent effect of oil prices on production exists.  However modest lagged effects of between 2 to 5 years exists.  These results are in line with the intuition

@@

@@

@@

A cleaned data set and the full code for the analysis are available at my website \url{jmaurit.github.io#oil_prices}.

@@

\begin{figure}
	\includegraphics[width=1\textwidth]{figures/mc_plot.png}
	\caption{Results from a Monte-Carlo experiment. I generate data mimicking oil price production from a set of fields, as displayed in the top frame. Using GLM and GAM models, and regenerating the error component in the data, I replicate estimates of the effect of price on production, which is set at $\beta = 0$ and $\beta=0.05$. The density of the results for the GLM model are shown in the lower left panel, where a clear and large bias is shown relative to the true $\beta$ which are represented by the vertical lines.  A GAM model however provides estimates close to the true $\beta$}
	\label{mc_results}
\end{figure}

@@

@@

@@

\section{software}
I use the R statistical programming package for all the analysis in this article \citep{r_core_team_r:_2013}.  I use the R package ggplot2 for plotting \citep{wickham_ggplot2:_2009}, plyr for data manipulation and cleaning \citep{wickham_split-apply-combine_2011}, texreg for table formatting \citep{leifeld_texreg:_2013} and mgcv for implementation of the Generalized Additive Models \citep{wood_fast_2011}.

@@

% \end{spacing}
% \end{document} for shorter version with no appendix
\FloatBarrier
\end{document}
